Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model

### 2015-04-24 msc-publication.tex
add streamfunction maths, model equations

%but are known to suffer from horizontal pressure gradient errors near steep terrain.
\abstract{Terrain following coordinates are widely used in operational models but the cut cell method has been proposed as an alternative that can more accurately represent the orography and associated atmospheric dynamics.  Because the type of grid is usually chosen during model implementation, it is typically necessary to use different models in order to compare the accuracy of different grids.  On the contrary, we use a single C-grid finite volume model to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test of a stably stratified atmosphere at rest, and orographically induced gravity waves.  In addition, we formulate a new tracer advection test having a velocity field that is everywhere tangential to the terrain following coordinate surfaces.
The results of the advection tests demonstrate that tracer accuracy depends upon alignment of the flow with the grid.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature.  However, we find that the Lorenz computational mode is excited as a vertical zig--zag in potential temperature on the cut cell grid.
\TODO{this last sentence is weak: Although errors do vary between grids, overall, a high degree of accuracy is obtained for all tests on all grids.}}

@@

Several studies have found that cut cells produce more accurate results when compared to TF coordinates.  Spurious winds seen in TF coordinates are not present and errors do not increase with steeper terrain \citep{good2014}.  \TODO{any other studies?}

This study uses the nonhydrostatic model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between terrain following and cut cell grids for idealised, two-dimensional test cases from the literature.  Section 2 presents the formulation of the terrain following and cut cell grids used in the experiments that follow.  Section 3 analyses the results from two tracer advection tests, a test of a stably stratified atmosphere at rest, and orographically induced gravity waves.  Concluding remarks are made in section 4.

@@

\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h \label{eqn:btf}
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.  An example is shown in figure~\ref{fig:grid}a.

The sigma coordinate transform of \citet{phillips1957} is equivalent to the BTF coordinate transform since they both decay linearly.  However, since they decay with pressure rather than height, sigma coordinates also change with horizontal variations in pressure.

@@

@@

A series of standard, two-dimensional tests are performed over idealised orography.  For each test, results on the BTF, SLEVE and cut cell grid are compared.  The first test from \citet{schaer2002} advects a tracer in a horizontal velocity field.  Second, a new tracer advection test is formulated employing a terrain following velocity field to challenge the advection scheme on more orthogonal grids.  The third test solves the Euler equations for a stably stratified atmosphere at rest, following \citet{klemp2011}.  Finally, as specified by \citet{schaer2002}, a test of orographically-induced gravity waves is performed.

We use a finite volume discretisation of the fully-compressible Euler equations described by \citet{weller-shahrokhi2014}.  The model has a curl-free pressure gradient formulation, an upwind-biased cubic advection scheme, and uses a Lorenz staggering of thermodynamic variables.  No explicit diffusion is used in any of the tests.

@@

The domain is \SI{300}{\kilo\meter} wide and \SI{25}{\kilo\meter} high, discretized onto a grid with \(\Delta x = \SI{1}{\kilo\meter}\) and \(\Delta z^\star = \SI{500}{\meter}\).  The terrain is wave-shaped, specified by the surface height, \(h\), such that
\begin{subequations}
\begin{align}
   h(x) &= h^\star \cos^2 ( \alpha x )
%
\intertext{where}
%
   h^\star(x) &= \left\{ \begin{array}{l l}
       h_0 \cos^2 ( \beta x ) & \quad \text{if $| x | < a$} \\
	0 & \quad \text{otherwise}
    \end{array} \right.
\end{align}
\end{subequations}
where $a = \SI{25}{\kilo\meter}$ is the mountain half-width, $h_0 = \SI{3}{\kilo\meter}$ is the maximum mountain height, $\lambda = \SI{8}{\kilo\meter}$ is the wavelength, \(\alpha = \pi / \lambda\) and \(\beta = \pi / 2a\).  On the SLEVE grid, the large-scale component $h_1$ is given by \(h_1(x) = h^\star(x) / 2\)
and $s_1 = \SI{15}{\kilo\meter}$ is the large scale height, and $s_2 = \SI{2.5}{\kilo\meter}$ is the small scale height.  The optimisation of SLEVE by \citet{leuenberger2010} is not used, so the exponent $n = 1$.

The wind is entirely horizontal and is prescribed as
\begin{align}
	u(z) = u_0 \left\{ \begin{array}{l l}
		1 & \quad \text{if $z \geq z_2$} \\
		\sin^2 \left( \frac{\pi}{2} \frac{z - z_1}{z_2 - z_1} \right) & \quad \text{if $z_1 < z < z_2$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.	
\end{align}
where $u_0 = \SI{10}{\meter\per\second}$, $z_1 = \SI{4}{\kilo\meter}$ and $z_2 = \SI{5}{\kilo\meter}$.
This results in a constant wind aloft, and zero flow at \SI{4}{\kilo\meter} and below.  To ensure the discrete velocity field is non-divergent, velocities are prescribed at cell faces by summing the streamfunction at the centre of all face edges.

@@

Tracer magnitude and shape are well-preserved on all grids, both above the mountain at \(t = \SI{5000}{\second}\) and past the mountain at \(t = \SI{10000}{\second}\).  Advection is most accurate on the cut cell grid (figure~\ref{fig:advection-tracer}c) and regular grid (not shown).  As found by \citet{good2014}, the result is the same on both grids.  This is to be expected since the wind is zero in the region of the ground and flow aloft is aligned with the grids.  On the BTF grid, the tracer is less distorted by the cubic upwind-biased scheme (figure~\ref{fig:advection-tracer}b) compared to the fourth-order centred scheme from \citet{schaer2002} (figure~\ref{fig:advection-tracer}a).

@@

At $t = \SI{10000}{\second}$ the tracer has passed over the mountain.  The horizontal position of the tracer centre can be calculated by integrating along the trajectory to find $t$, the time taken to pass from one side of the mountain to the other:
\begin{multline}
t = \frac{x}{u_0} - \frac{h_0}{16 u_0 H} \left[ 4x + \frac{\sin 2 (\alpha + \beta) x}{\alpha + \beta} \right.+ \\
\left. \frac{\sin 2(\alpha - \beta) x}{\alpha - \beta} + 2 \left( \frac{\sin 2\alpha x}{\alpha} + \frac{\sin 2\beta x}{\beta} \right) \right]
\end{multline}
Hence, we find that \(x(t=\SI{10000}{\second}) = \SI{51577.4}{\meter}\).  Because the velocity field is non-divergent, the flow accelerates over mountain ridges and the tracer travels \SI{1577.4}{\meter} further compared to advection in the horizontal velocity field.  Tracer height is unchanged because advection is along the terrain following coordinate surface.

@@

\subsection{Resting atmosphere}
\label{sec:resting}

@@

Sponge layers are added to the upper \SI{10}{\kilo\meter} and leftmost \SI{10}{\kilo\meter} at the inlet boundary to damp the reflection of waves.
The term $\mu \rho \bf{u}$ is subtracted from the momentum equation \TODO{include the equation set in this paper, or cite weller-shahrokhi2014 here?} where the damping function $\mu$ is adapted from \citet{melvin2010} such that
\begin{align}
	\mu(x, z) &= \mu_\mathrm{upper} + \mu_\mathrm{inlet} \\
	\mu_\mathrm{upper}(z) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{z - z_B}{H - z_B} \right) & \text{if } z \geq z_B \\
		0 & \text{otherwise} \\
	\end{cases} \\
	\mu_\mathrm{inlet}(x) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{x_I - x}{x_I - x_0} \right) & \text{if } x < x_I \\
		0 & \text{otherwise}
	\end{cases}
\end{align}
where $\overline{\mu} = 1.2$ is the damping coefficient, $z_B = \SI{20}{\kilo\meter}$ is the bottom of the sponge layer, $H = \SI{30}{\kilo\meter}$ is the top of the domain, $x_0 = \SI{-150}{\kilo\meter}$ is the leftmost limit of the domain and $x_I = \SI{-140}{\kilo\meter}$ is the rightmost extent of the inlet sponge layer.  The sponge layer is only active on faces whose normal is vertical so that it damps vertical momentum only. \TODO{is this true of the inlet sponge as well?}

@@

The model by \citet{weller-shahrokhi2014} has a Lorenz staggering of variables such that $\theta$ is stored at cell centres and velocity at cell faces.  The inaccuracies in potential temperature become invisible when calculating momentum because $\theta$ is interpolated onto cell faces, so discrete hydrostatic balance is maintained.  This is a manifestation of the Lorenz computational mode, which is a disadvantage of the Lorenz grid that has been widely discussed in the literature \citep{arakawa-konor1996,holdaway2013b}.  In models that include moist processes, the Lorenz computational mode can disrupt clouds and generate spurious precipitation \citep{hollingsworth1995}.

To summarize, results of the gravity waves test on all grids are in good agreement with the reference solution from \citet{melvin2010}.  The most prominent errors are found only on the cut cell grid, where the potential temperature errors near the ground are maintained by the Lorenz computational mode.

@@

Across all tests, a high degree of accuracy was achieved for all grids.  Even on the highly-distorted BTF grid, which is often found to give poor results \citet{schaer2002,klemp2011,good2014}, errors were often small in the tests presented here.  In the first two tests, tracers were advected in horizontal and terrain following velocity fields.  We found that the accuracy of the upwind-biased cubic advection scheme depended upon alignment of the flow with the grid.  Spurious vertical velocities were small in the resting atmosphere test, reaching a maximum of $\sim \SI{0.35}{\meter\per\second}$ on the BTF grid, compared to a maximum of $\sim \SI{10}{\meter\per\second}$ found by \citet{klemp2011}.  In the gravity waves test, vertical velocities were in good agreement with the reference solution from \citet{melvin2010} across all grids.

@@

\acknowledgments
Start acknowledgments here.  \TODO{John, Terry.  Assuming only Hilary is co-author...?}

>>>>

\abstract{Terrain following coordinates are widely used in operational models but the cut cell method has been proposed as an alternative that can more accurately represent the orography and associated atmospheric dynamics.  Because the type of grid is usually chosen during model implementation, it is typically necessary to use different models in order to compare the accuracy of different grids.  On the contrary, we use a single C-grid finite volume model to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test of a stably stratified atmosphere at rest, and orographically induced gravity waves.  In addition, we formulate a new tracer advection test having a velocity field that is everywhere tangential to the terrain following coordinate surfaces.
The results of the advection tests demonstrate that tracer accuracy depends upon alignment of the flow with the grid.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature.  However, we find that the Lorenz computational mode is excited as a vertical zig--zag in potential temperature on the cut cell grid.
\TODO{this last sentence is weak: Although errors do vary between grids, overall, a high degree of accuracy is obtained for all tests on all grids.}}

@@

Several studies have shown examples where cut cells produce more accurate results when compared to TF coordinates.  Spurious winds seen in TF coordinates are not present and errors do not increase with steeper terrain \citep{good2014}.  A comparison of TF and cut cells using real initial data by \citet{steppeler2013} found that precipitation and wind forecasts were more accurately in the cut cell model.

This study uses the nonhydrostatic model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between terrain following and cut cell grids for idealised, two-dimensional test cases from the literature.  Section 2 presents the formulation of the terrain following and cut cell grids used in the experiments that follow.  \TODO{section 3} Section 4 analyses the results from two tracer advection tests, a test of a stably stratified atmosphere initially at rest, and orographically induced gravity waves.  Concluding remarks are made in section 5.

@@

\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h \label{eqn:btf}
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the physical height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.  An example is shown in figure~\ref{fig:grid}a.

The sigma coordinate transform of \citet{phillips1957} is equivalent to the BTF coordinate transform since they both decay linearly.  However, because \(\sigma = p/p_s\) decays with pressure rather than height, sigma coordinates also vary with surface pressure, \(p_s(x, t)\).

@@

\section{Model equations}
\label{sec:model}

The C-grid finite volume model from \citet{weller-shahrokhi2014} discretises the fully-compressible Euler equations, given by
\begin{subequations}
\begin{align}
	\text{Momentum} &\ &\  	\frac{\partial \rho \bm{u}}{\partial t} + \nabla \cdot \rho \bm{uu} &= \rho \bm{g} - c_p \rho \theta \nabla \Pi \label{eq:momentum} \\
	\text{Continuity} &\ &\	\frac{\partial \rho}{\partial t} + \nabla \cdot \rho \bm{u} &= 0 \\
	\text{$\theta$, flux form} &\ &\ \frac{\partial \rho \theta}{\partial t} + \nabla \cdot \rho \bm{u} \theta &= 0 \\
	\text{$\theta$, advective form} &\ &\ \frac{\partial \theta}{\partial t} + \bm{u} \cdot \nabla \theta &= 0 \\
	\text{Equation of state} &\ &\ \Pi^{(1 - \kappa)/\kappa} &= \frac{R \rho \theta}{p_0}
\end{align}
\end{subequations}
where \(\rho\) is the density, \(\bm{u}\) is the velocity field, \(\bm{g}\) is the gravitational acceleration, \(c_p\) is the heat capacity at constant pressure, \(\theta = T \left(p_0/p\right)^\kappa\) is the potential temperature, \(T\) is the tempereature, \(p\) is the pressure, \(p_0\) is a reference pressure, \(\Pi = \left(p / p_0 \right)^\kappa\) is the Exner function of pressure, and \(\kappa = R/c_p\) is the gas constant to heat capacity ratio.

@@

A series of two-dimensional tests are performed over idealised orography.  For each test, results on the BTF, SLEVE and cut cell grid are compared.  The first test from \citet{schaer2002} advects a tracer in a horizontal velocity field.  Second, a new tracer advection test is formulated employing a terrain following velocity field to challenge the advection scheme on more orthogonal grids.  The third test solves the Euler equations for a stably stratified atmosphere initially at rest, following \citet{klemp2011}.  Finally, as specified by \citet{schaer2002}, a test of orographically-induced gravity waves is performed.

We use a finite volume discretisation of the fully-compressible Euler equations described by \citet{weller-shahrokhi2014}.  The model has a curl-free pressure gradient formulation and an upwind-biased cubic advection scheme.  It uses a Lorenz staggering of thermodynamic variables such that $\theta$ is stored at cell centres and velocity at cell faces.  No explicit diffusion is used in any of the tests.

@@

The domain is \SI{300}{\kilo\meter} wide and \SI{25}{\kilo\meter} high, discretized onto a grid with \(\Delta x = \SI{1}{\kilo\meter}\) and \(\Delta z^\star = \SI{500}{\meter}\).  The terrain is wave-shaped, specified by the surface height, \(h\), such that
\begin{subequations}
\begin{align}
   h(x) &= h^\star \cos^2 ( \alpha x )
%
\intertext{where}
%
   h^\star(x) &= \left\{ \begin{array}{l l}
       h_0 \cos^2 ( \beta x ) & \quad \text{if $| x | < a$} \\
	0 & \quad \text{otherwise}
    \end{array} \right.
\end{align}
\end{subequations}
where $a = \SI{25}{\kilo\meter}$ is the mountain envelope half-width, $h_0 = \SI{3}{\kilo\meter}$ is the maximum mountain height, $\lambda = \SI{8}{\kilo\meter}$ is the wavelength, \(\alpha = \pi / \lambda\) and \(\beta = \pi / 2a\).  On the SLEVE grid, the large-scale component $h_1$ is given by \(h_1(x) = h^\star(x) / 2\)
and $s_1 = \SI{15}{\kilo\meter}$ is the large scale height, and $s_2 = \SI{2.5}{\kilo\meter}$ is the small scale height.  The optimisation of SLEVE by \citet{leuenberger2010} is not used, so the exponent $n = 1$.

The wind is entirely horizontal and is prescribed as
\begin{align}
	u(z) = u_0 \left\{ \begin{array}{l l}
		1 & \quad \text{if $z \geq z_2$} \\
		\sin^2 \left( \frac{\pi}{2} \frac{z - z_1}{z_2 - z_1} \right) & \quad \text{if $z_1 < z < z_2$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.	
\end{align}
where $u_0 = \SI{10}{\meter\per\second}$, $z_1 = \SI{4}{\kilo\meter}$ and $z_2 = \SI{5}{\kilo\meter}$.
This results in a constant wind aloft, and zero flow at \SI{4}{\kilo\meter} and below.

To ensure the discrete velocity field is non-divergent, velocities are prescribed at cell faces by summing the streamfunction, \(\Psi\), at the centre of all face edges.  Since \(u = \partial \Psi / \partial z\), the streamfunction is found by vertical integration of the velocity profile:
\begin{align}
	\Psi(z) &= \frac{u_0}{2} \left\{ \begin{array}{l l}
		\left( 2z - z_1 - z_2 \right) & \enskip \text{if $z > z_2$} \\
		z - z_1 - \frac{z_2 - z_1}{\pi} \sin \left(\pi \frac{z - z_1}{z_2-z_1}\right) & \enskip \text{if $z_1 < z \leq z_2$} \\
		0 & \enskip \text{if $z \leq z_1$}
	\end{array} \right.
\end{align}

@@

Tracer magnitude and shape are well-preserved on all grids, both above the mountain at \(t = \SI{5000}{\second}\) and past the mountain at \(t = \SI{10000}{\second}\).  In this test, advection is most accurate on the cut cell grid (figure~\ref{fig:advection-tracer}c) and regular grid (not shown).  As found by \citet{good2014}, the result is the same on both grids.  This is to be expected since the wind is zero in the region of the ground and flow aloft is aligned with the grids.  On the BTF grid, the tracer is less distorted by the cubic upwind-biased scheme (figure~\ref{fig:advection-tracer}b) compared to the fourth-order centred scheme from \citet{schaer2002} (figure~\ref{fig:advection-tracer}a).

@@

At $t = \SI{10000}{\second}$ the tracer has passed over the mountain.  The horizontal position of the tracer centre can be calculated by integrating along the trajectory to find $t$, the time taken to pass from one side of the mountain to the other:
\begin{multline}
t = \frac{x}{u_0} - \frac{h_0}{16 u_0 H} \left[ 4x + \frac{\sin 2 (\alpha + \beta) x}{\alpha + \beta} \right.+ \\
\left. \frac{\sin 2(\alpha - \beta) x}{\alpha - \beta} + 2 \left( \frac{\sin 2\alpha x}{\alpha} + \frac{\sin 2\beta x}{\beta} \right) \right]
\end{multline}
Hence, we find that \(x(t=\SI{10000}{\second}) = \SI{51577.4}{\meter}\).  Because the velocity field is non-divergent, the flow accelerates over mountain ridges and the tracer travels \SI{1577.4}{\meter} further compared to advection in the purely horizontal velocity field.  Tracer height is unchanged downwind of the mountains because advection is along the terrain following coordinate surface.

@@

\subsection{Stratified atmosphere initially at rest}
\label{sec:resting}

@@

Sponge layers are added to the upper \SI{10}{\kilo\meter} and leftmost \SI{10}{\kilo\meter} at the inlet boundary to damp the reflection of waves.
The term $\mu \rho \bm{u}$ is subtracted from the momentum (equation \ref{eq:momentum}) where the damping function, \(\mu\), is adapted from \citet{melvin2010} such that
\begin{align}
	\mu(x, z) &= \mu_\mathrm{upper} + \mu_\mathrm{inlet} \\
	\mu_\mathrm{upper}(z) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{z - z_B}{H - z_B} \right) & \text{if } z \geq z_B \\
		0 & \text{otherwise} \\
	\end{cases} \\
	\mu_\mathrm{inlet}(x) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{x_I - x}{x_I - x_0} \right) & \text{if } x < x_I \\
		0 & \text{otherwise}
	\end{cases}
\end{align}
where $\overline{\mu} = 1.2$ is the damping coefficient, $z_B = \SI{20}{\kilo\meter}$ is the bottom of the sponge layer, $H = \SI{30}{\kilo\meter}$ is the top of the domain, $x_0 = \SI{-150}{\kilo\meter}$ is the leftmost limit of the domain and $x_I = \SI{-140}{\kilo\meter}$ is the rightmost extent of the inlet sponge layer.  The sponge layer is only active on faces whose normal is vertical so that it damps vertical momentum only.

@@

The inaccuracies in potential temperature become invisible when calculating momentum because $\theta$ is interpolated onto cell faces, so discrete hydrostatic balance is maintained.  This is a manifestation of the Lorenz computational mode, which is a disadvantage of the Lorenz grid that has been widely discussed in the literature \citep{arakawa-konor1996,holdaway2013b}.  In models that include moist processes, the Lorenz computational mode can disrupt clouds and generate spurious precipitation \citep{hollingsworth1995}.

To summarize, results of the gravity waves test on all grids are in good agreement with the reference solution from \citet{melvin2010}.  The most prominent errors are found only on the cut cell grid, where the potential temperature errors near the ground result from exciting the Lorenz computational mode.

@@

Across all tests, a high degree of accuracy was achieved for all grids.  Even on the highly-distorted BTF grid, which have previously been found to give poor results \citep{schaer2002,klemp2011,good2014}, errors were often small in the tests presented here.  In the first two tests, tracers were advected in horizontal and terrain following velocity fields.  We found that the accuracy of the upwind-biased cubic advection scheme depended upon alignment of the flow with the grid.

Spurious vertical velocities were small in the resting atmosphere test, reaching a maximum of $\sim \SI{0.35}{\meter\per\second}$ on the BTF grid, compared to a maximum of $\sim \SI{10}{\meter\per\second}$ found by \citet{klemp2011}.  In the gravity waves test, vertical velocities were in good agreement with the reference solution from \citet{melvin2010} across all grids.

@@

\acknowledgments
Start acknowledgments here.  \TODO{John, Terry.  NERC grant NE/H015698/1.  Assuming only Hilary is co-author...?}
